---
title: "Stat_models_project"
output: html_document
date: "2022-11-28"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Checking out the data for the missing values:

```{r}
data <- read.table('Smoker_Epigenetic_df.csv', header=T, sep = ",")

apply(is.na(data), 2, sum)
apply(is.na(data), 2, which)

```
As we can see we have 62 samples where all data on methylation is missing. So we can just drop these values.

```{r}
data <- data[-which(is.na(data$cg03706273)), ]
data <- data[, -c(1)]
head(data, n=3)
```



```{r, echo=FALSE}
library(tidyverse)
library(dplyr)

ggpairs(data, columns = 1:5)

```

```{r}
data_m <- melt(data, na.rm = FALSE, id = names(data)[seq(1,4, by=1 )])
head(data_m, n=3)
```

```{r, fig.width=18, fig.height=4}

ggplot(data_m, aes(factor(variable), value, fill=factor(variable))) + geom_violin()
```

We can see that the majority of islands have a bimodal distribution with higher density around 0.5 and around either 0.8 or 0.1.

Now we can build the barplots for each island to compare the methylation level between smokers and non-smokers:

```{r, fig.width=18, fig.height=4}
ggplot(data_m, aes(factor(variable), value, color = Smoking.Status)) +  # ggplot function
  geom_boxplot()

```

We can observe that the difference between smoking and non-smoking people is so that usually only mean stays the same but quartiles are changing. Thus for non-smokers we have more narrow distributions in most cases comparing to the smokers.

Now when we understood the distribution of data. The better fit to ...

\textcolor{red}{compare vars instead of raw values bc they should really differ} 

```{r}


```

First, we can calculate the deviance from mean for each row for each island:

```{r}
data_diff <- data
for(col in colnames(data)[4:length(names(data_diff))]){
  data_diff[[col]] <- data_diff[[col]] - mean(data_diff[[col]])
}

data_diff$mean_per_row <- rowMeans(data_diff[, 4:length(names(data_diff))])

'''
NOT WORKING BUT NICE IDEA

# we can now code the means by calculating the confidence intervals
significant_diff <- c(mean(data_diff$mean_per_row) - 3*sd(data_diff$mean_per_row), 
                      mean(data_diff$mean_per_row) + 3*sd(data_diff$mean_per_row))


data_diff$mean_cat <- ifelse((data_diff$mean_per_row > significant_diff[1]) & (data_diff$mean_per_row < significant_diff[2]), 0, 1)

EVERYTHING IS IN THE INTERVAL
'''

data_diff$mean_cat <- abs(data_diff$mean_per_row)

# taking abs and coding as 1 if abs is more than average and as 0 is less
data_diff$mean_cat <- ifelse(data_diff$mean_cat > mean(data_diff$mean_cat), 1, 0)

head(data_diff, n = 3)
```

```{r}
table(data_diff$mean_cat)
```


Then we can try to construct some models based on this columns and then compare its performance on other columns:

```{r}
#install.packages("Correlplot")
library(Correlplot)
attach(data_diff)

model_log1 <- glm(mean_cat~Smoking.Status, 
              family = binomial(link = 'logit'), # to use logistic regression
              trace=TRUE)

summary(model_log1)

```

```{r}
anova(model_log1,test="Chisq") 
```


```{r, eco=False}

plot(data_diff$Smoking.Status, data_diff$mean_cat)
     #, xlab="Last weight mother", ylab="LBW", ylim=c(-0.1,1.1))

# type - not logit
curve(predict(model_log1, data.frame(data_diff$Smoking.Status=x), type="resp"),add=TRUE,col="blue",lwd=2)
points(data_diff$Smoking.Status, fitted(model_log1), pch=1, col="blue") # fitted probabilities

```


then:

1) improve this model
2) make linear model and compare
3) perform on all cpg islands


```{r}

```


```{r}


```

