---
title: "Stat_models_project"
output: html_document
date: "2022-11-28"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Checking out the data for the missing values:

```{r}
data <- read.table('Smoker_Epigenetic_df.csv', header=T, sep = ",")

apply(is.na(data), 2, sum)
apply(is.na(data), 2, which)

```
As we can see we have 62 samples where all data on methylation is missing. So we can just drop these values.

```{r}
data <- data[-which(is.na(data$cg03706273)), ]
data <- data[, -c(1)]
tail(data, n=3)
```

```{r}
data$Gender<-sapply(data$Gender, tolower)
table(data$Gender)    

ggplot(data, aes(x=Age, color=Gender)) +
  geom_histogram(fill="white", alpha=0.5, position="identity")
```



```{r, echo=FALSE}
library(tidyverse)
library(dplyr)

ggpairs(data, columns = 1:5)

```

```{r}
data_m <- melt(data, na.rm = FALSE, id = names(data)[seq(1,4, by=1 )])
head(data_m, n=3)
```

```{r, fig.width=18, fig.height=4}

ggplot(data_m, aes(factor(variable), value, fill=factor(variable))) + geom_violin()
```

We can see that the majority of islands have a bimodal distribution with higher density around 0.5 and around either 0.8 or 0.1.

Now we can build the barplots for each island to compare the methylation level between smokers and non-smokers:

```{r, fig.width=18, fig.height=4}
ggplot(data_m, aes(factor(variable), value, color = Smoking.Status)) +  # ggplot function
  geom_boxplot()

```

We can observe that the difference between smoking and non-smoking people is so that usually only mean stays the same but quartiles are changing. Thus for non-smokers we have more narrow distributions in most cases comparing to the smokers.


```{r}
# making new dataframe for the corrected values

data_diff <- data.frame(data) 
head(data_diff, 3)
```

```{r}
data_diff_only_nonsm <- data_diff[data_diff$Smoking.Status=="never", ]

# the vector of means for each island for nonsmokers
means_nonsmokers_vector <- append(c("never", "fm"), colMeans(data_diff_only_nonsm[sapply(data_diff_only_nonsm[], is.numeric)]))
means_nonsmokers_vector
```

```{r}
# calculating the deviance from the nonsmoker-mean for each parient
for(col in colnames(data)[5:length(names(data_diff))]){
  data_diff[[col]] <- data_diff[[col]] - as.numeric(means_nonsmokers_vector[col])
}

# calculating average deviance for each parient along all columns
data_diff$mean_per_row <- rowMeans(data_diff[, 4:length(names(data_diff))])

head(data_diff, 3)

```
```{r}
# we can have a look at the distribution of the mean deviation - it shoud be around zero
library(ggplot2)
ggplot(data_diff, aes(x=mean_per_row)) +  geom_histogram()
```
     

```{r}
# now we can take absolute value and code the column
data_diff$mean_cat <- abs(data_diff$mean_per_row)

# taking abs and coding as 1 if abs is more than mean(=0) + std and as 0 if it's is less
data_diff$mean_cat <- ifelse(data_diff$mean_cat > sd(data_diff$mean_cat), 1, 0)

head(data_diff, n = 3)
```


```{r}
# now we can construct a dataset with all CpG islands coded as 1 or 0

data_cat_all_cpg <- data.frame(data_diff)[-c(621), ]

for(col in colnames(data_cat_all_cpg)[4:length(names(data_cat_all_cpg))]){
  data_cat_all_cpg[[col]] <- ifelse(data_cat_all_cpg[[col]] > sd(data_cat_all_cpg[[col]]), 1, 0)
}
```

```{r}
table(data_diff$mean_cat)
```

```{r}
data_diff$Smoking.Status <- ifelse(data_diff$Smoking.Status == "never", 0, 1)
data_cat_all_cpg$Smoking.Status <- ifelse(data_cat_all_cpg$Smoking.Status == "never", 0, 1)
head(data_cat_all_cpg, 3)
```

## MODELLING

Then we can try to construct some models based on this columns and then compare its performance on other columns:

```{r, error=FALSE}
#install.packages("Correlplot")
library(Correlplot)

# model for the generalized column ONLY SMOKING STATUS
model_log1 <- glm(data_diff$mean_cat~data_diff$Smoking.Status, 
              family = binomial(link = 'logit'), # to use logistic regression
              trace=TRUE)

summary(model_log1)

```
```{r, echo=False}
attach(data_diff)

plot(jitter(data_diff$Smoking.Status, amount = 0.1), jitter(data_diff$mean_cat, amount = 0.1), xlab="Smoking status",
     ylab="Methylation", xlim=c(-0.1,1.1))

curve(predict(model_log1, data.frame(Smoking.Status = x), type="resp"), add=TRUE,col="blue",lwd=2)

points(data_diff$Smoking.Status, fitted(model_log1), pch=1, col="blue") # fitted probabilities

```


```{r}
# model for the generalized column ONLY SMOKING STATUS + AGE + GENDER

model_log_sag <- glm(data_diff$mean_cat~data_diff$Smoking.Status + data_diff$Age + data_diff$Gender, 
              family = binomial(link = 'logit'), # to use logistic regression
              trace=TRUE)

summary(model_log_sag)
```

```{r}

# model for the generalized column ONLY SMOKING STATUS + GENDER

model_log_sg <- glm(data_diff$mean_cat~data_diff$Smoking.Status + data_diff$Gender, 
              family = binomial(link = 'logit'), # to use logistic regression
              trace=TRUE)

summary(model_log_sg)
```

```{r}
# model for the generalized column ONLY SMOKING STATUS + AGE

model_log_sa <- glm(data_diff$mean_cat~data_diff$Smoking.Status + data_diff$Age, 
              family = binomial(link = 'logit'), # to use logistic regression
              trace=TRUE)

summary(model_log_sa)
```

## MODELS FOR THE ISLANDS

```{r}

# model for the first island
model_log_cg00214611 <- glm(data_cat_all_cpg$mean_cat~data_cat_all_cpg$Smoking.Status, 
              family = binomial(link = 'logit'), # to use logistic regression
              trace=TRUE)

summary(model_log_cg00214611)
```


```{r}
all_islands <- names(data_cat_all_cpg)[4:(length(names(data_cat_all_cpg))-2)]

for (island in all_islands) {
  model = glm(data_cat_all_cpg[[island]] ~ data_cat_all_cpg$Smoking.Status + data_cat_all_cpg$Gender, family = binomial(link = "logit"), trace = F )
  print(island)
  print(summary(model))
  print("----------------------------------------------------------------------------")
}

```

```{r}


```

```{r}


```

```{r}


```

```{r}


```   

